---
title: "Connection and Database Management"
output:
  html_document:
    toc: true 
    toc_depth: 3
    number_sections: false
    theme: united
vignette: >
  %\VignetteIndexEntry{Connection and Database Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 6L, tibble.print_max = 6L, digits = 3)
```

## Connection overview
Connecting to Tereadata Vantage can be performed using the `td_create_context` function or the `DBI::dbConnect` function. Once the connection is established, a `context` must be initialized. The results of tdplyr analytic functions are stored as nonperistent tables in the SQL Engine database. The context allows the user specify a database to store these nonpersistent tables. The process of initializing a context can either be performed using the `td_create_context` function or the `td_set_context` function.

The `td_remove_context` function can be used to disconnect the connection and remove the context. Removing the context triggers the removal of all nonpersistent work tables, a process also known as garbage collection.

### Supported drivers
TeradataNative Driver (>= 16.20.0.18)

Teradata ODBC driver (>= 16.20) - Use of tdplyr with Teradata ODBC Driver will be deprecated.

## Use td_create_context to connect and initialize the context
The function `td_create_context` establishes the connection and initializes the context. There are two ways of creating connection. 
```{r, results = "hide", warning = FALSE, message = FALSE}
library(tdplyr)
```

1. Create context using  Teradata ODBC Driver
```{r odbc, results = 'hide', eval=FALSE}
con <- td_create_context("TeradataDSN")
```


2. Create context using Teradata SQL driver for R
```{r, echo=FALSE, include=FALSE}
# Reading Vantage cluster details from td_config.cfg
config_file <- system.file("extdata", "td_config.cfg", package = "tdplyr")
config_df <- read.csv(config_file, sep = ":", stringsAsFactors = FALSE, header = TRUE)
config_list <- unlist(lapply(config_df$value, function(x) {trimws(x)}))
host_name <- config_list[1]
user_name <- config_list[2]
password  <- config_list[3]
```

```{r native, results = 'hide'}
# host_name = Fully qualified domain name or IP address of the Teradata Vantage System.
# user_name = UserId for the connection.
# password  = Password for the connection.
con <- td_create_context(host = host_name, uid = user_name, pwd = password, dType = "native")
```

The class of the connection object is same for both the drivers.
```{r}
class(con)[1] # The connection object is an object of class: "Teradata"
```

### Use td_get_context to retrieve attributes of the current context

The `td_get_context` function returns a named list of attributes in the current context.

```{r}
names(td_get_context())
```

The connection object in the current context can also be retrieved using the `td_get_context` function.
```{r}
con <- td_get_context()$connection
```

### Use td_remove_context to disconnect the connection and remove the context
The function `td_remove_context` can be used to disconnect the connection and remove the context. Removing the context triggers the removal of all nonpersistent work tables, a process also known as garbage collection.
```{r}
td_remove_context()
```

### Specify a temporary database to store tables created by analytic functions
Analytic functions need to create temporary database objects (table / view) for processing data. The database where these objects are created can be specified using the `temp.database` argument in the `td_create_context` function. If the `temp.database` argument is not specified, then the default database of the connection is used.

1. Specifying temporary database using Teradata ODBC driver.
```{r, eval=FALSE}
td_create_context("TeradataDSN", temp.database = "DATABASE_1")
```

2. Specifying temporary database using Teradata SQL driver for R.
```{r}
con <- td_create_context(host = host_name, uid = user_name, pwd = password, dType = "native", temp.database = "DATABASE_1")
td_get_context()$temp.database
```

```{r}
td_remove_context()
```

## Use td_set_context to initialize the context
If a connection is established using the `DBI::dbConnect` function, then use the `td_set_context` function to initialize the context.

1. Set context using Teradata ODBC driver.
```{r, eval=FALSE}
con <- DBI::dbConnect(odbc::odbc(), dsn = "TeradataDSN")
td_set_context(con)
```

2. Set context using Teradata SQL driver for R.
```{r, eval=TRUE}
con <- DBI::dbConnect(tdplyr::NativeDriver(), host = host_name, uid = user_name, pwd = password, dType = "native")
td_set_context(con)
```

If the `temp.database` argument is not specified in the `td_set_context` function, then the default database of the supplied connection object is used as the temporary database.

The function `td_set_context` can also be used to change the temporary database of the current context.

```{r}
td_set_context(con, "DATABASE_1")
td_get_context()$temp.database
```

```{r, echo=FALSE}
td_remove_context()
```


### Change the temporary database of an existing context
The temporary database in the current context can be changed using the `td_set_context` function without having to creating a new context.

1. Changing temporary database using Teradata ODBC driver.
```{r, eval=FALSE}
td_create_context("TeradataDSN", temp.database = "DATABASE_1")

# Change the temporary database in the current context
td_set_context(td_get_context()$connection, "DATABASE_2")
```

2. Changing temporary database using Teradata SQL driver for R.
```{r}
con <- td_create_context(host = host_name, uid = user_name, pwd = password, dType = "native", temp.database = "DATABASE_1")
td_get_context()$temp.database

# Change the temporary database in the current context
td_set_context(td_get_context()$connection, "DATABASE_2")
td_get_context()$temp.database
```


## Database management functions
List ODBC data source names configured in your system.
```{r, results = "hide"}
odbc::odbcListDataSources()
```

View connection related information.
```{r, results = "hide"}
con <- td_get_context()$connection
DBI::dbGetInfo(con)
```
Alternatively, use the `db_desc` function to view the connection related information.
```{r, results = "hide"}
dplyr::db_desc(con)
```

The output of the `db_desc` function can be changed using the `td.db_desc.output` option. Set a named list containing `db`, `driver`, and `machine` to display the Advanced SQL Engine version, client driver version and the server machine name respectively.
```{r}
options(td.db_desc.output = list(db = TRUE, driver = TRUE, machine = FALSE))
dplyr::db_desc(con)
```

View only the volatile tables active in the given connection.
```{r}
# create a volatile table using the copy_to function
dplyr::copy_to(con, data.frame(col1 = c(1:3)), "volatile_table_1", temporary = TRUE)
# View all volatile tables active in the given connection
volatile_tables <- DBI::dbGetQuery(con, "help volatile table")
volatile_tables$"Table SQL Name"
```

View column names for a given table.
```{r, results = "hide"}
DBI::dbListFields(con, volatile_tables$"Table SQL Name")
```

Execute a SQL query and retrieve the result in an R native data frame.
```{r}
df <- DBI::dbGetQuery(con, "select database;")
df
```

```{r}
td_remove_context()
```

View all tables in the default database of the current context.
Note: This command also lists all the volatile (or temporary) tables active in the given connection. dbListTables is supported with ODBC connection only.

```{r , results = 'hide', eval=FALSE}
con1 <- td_create_context(dsn = "TeradataDSN")
tables <- DBI::dbListTables(con1, schema = td_get_context()$default.database, table_name = "%")
tables
```



