---
title: "Miscellaneous tdplyr functions"
output:
  html_document:
    toc: true 
    toc_depth: 3
    number_sections: false
    theme: united
vignette: >
  %\VignetteIndexEntry{Miscellaneous tdplyr functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 6L, tibble.print_max = 6L, digits = 3)
library(knitr)
library(dbplyr)
library(dplyr)
library(tdplyr)

opts <- list(db = FALSE, machine = FALSE, driver = FALSE)
options(td.db_desc.output = opts)

# Reading Vantage cluster details from td_config.cfg
config_file <- system.file("extdata", "td_config.cfg", package = "tdplyr")
config_df <- read.csv(config_file, sep = ":", stringsAsFactors = FALSE, header = TRUE)
config_list <- unlist(lapply(config_df$value, function(x) {trimws(x)}))

con <- td_create_context(host = config_list[1], uid = config_list[2], pwd = config_list[3], dType = "native")

```

# Introduction to tdplyr functions
This section introduces some useful tdplyr functions and their usage. Each function has a section in this vignette. Each section can be read in isolation, except where references are made to other sections or other vignettes. Each section also has a subsection called "Things to Note" for good practices and specific behavior to consider when using these tdplyr specific functions with Teradata Vantage.

Once the connection is established using **td_create_context**, you can load the example table `antiselect_input`and create tbl_teradata using the following statements:
```{r, results = "hide", warning=FALSE,  message = FALSE}
loadExampleData("antiselect_example", "antiselect_input")

# Create tbl_teradata object.
df <- tbl(con, "antiselect_input")

# Create a tibble from tbl_teradata object.
df1 <- as_tibble(df)

# Create a R dataframe from tbl_teradata object.
df2 <- as.data.frame(df)
```

## td_nrow
The function `td_nrow` retrieves number of rows in a tibble or data.frame object or tbl_teradata object. 

**Note:** The tbl_teradata objects are arranged to display ordered result while printing the output.
```{r}
# Print the tbl_teradata object.
df %>% arrange(rowids)

# Get number of rows of tbl_teradata object.
td_nrow(df)

# Get number of rows of tibble object.
td_nrow(df1)

# Get number of rows of R dataframe.
td_nrow(df2)
```
#### Notes
- The `nrow` function of R works for dataframes and tibble objects but not for tbl_teradata objects. So, to get the number of rows in a tbl_teradata object, you should use the function `td_nrow`.
- Please refer to dplyr function `tally` which is used to get number of the rows in R data.frames.

## td_sample
The function `td_sample` reduces the number of rows to be considered for further processing by returning one or more samples of rows. Sampling can be done in either of the three ways mentioned below:

*   Specifying list of numbers (number of rows in each sample)
*   Specifying list of fractions (proportion of the total number of rows in each sample)
*   Specifying list of numbers/fractions based on conditions (stratified sampling)

For more information about the function and its arguments, one can use the command `help(td_sample)` from R console.
```{r}
# Reducing the number of columns.
df <- df %>% select(rowids, orderid, priority, quantity)

# Example 1: Get two samples of 3 rows and 2 rows each. Note that the column 'sampleid' takes 2 values. Each sampleid value represents one sample.
df1 <- td_sample(df = df, n = c(3, 2))

# Print the results.
df1 %>% arrange(rowids)

# Example 2: Get a sample of 3 rows. Note that all the rows have sampleid = 1.
df1 <- td_sample(df = df, n = 3)

# Print the results.
df1 %>% arrange(rowids)

# Example 3: Get 50% of total rows.
df1 <- td_sample(df = df, n = 0.5)

# Print the results.
df1 %>% arrange(rowids)

# Example 4: Get two samples each containing 30% and 50% of rows.
df1 <- td_sample(df = df, n = c(0.3, 0.5))

# Print the results.
df1 %>% arrange(rowids)

# Example 5: Get 10 rows from a tbl_teradata object of 7 rows without setting with.replacement. Default value of the argument 'with.replacement' is FALSE. Note that only 7 rows are returned as the table doesn't contain 10 rows.
df1 <- td_sample(df = df, n = 10)

# Print the results.
df1 %>% arrange(rowids)

# Check the number of rows of 'df1'.
td_nrow(df1)

# Example 6: Get 10 rows from a tbl_teradata object of 7 rows setting with.replacement to TRUE. 'randomize = TRUE'ensures sampling is done across AMPs in large datasets. Note that 10 rows are returned and there are duplicates for some rows.
df1 <- td_sample(df = df, n = 10, with.replacement = TRUE, randomize = TRUE)

# Print the results.
df1 %>% arrange(rowids)

# Check the number of rows of 'df1'.
td_nrow(df1)

# Example 7 : Get 5 rows which satisfy the condition 'orderid < 300' from a tbl_teradata object. Here, only three rows are returned as the total number of rows which satisfy this condition is 3.
df1 <- td_sample(df, when_then = list("orderid < 300" = 5))

# Print the results.
df1 %>% arrange(rowids)

# Check the number of rows of 'df1'.
td_nrow(df1)

# If with.replacement is set to TRUE, then 5 rows will be returned.
df1 <- td_sample(df, when_then = list("orderid < 300" = 5), with.replacement = TRUE)

# Check the number of rows of 'df1'.
td_nrow(df1)

# Example 8: Get 4 rows (1 row in first sample and 3 rows in second sample) which satisfy the condition 'orderid < 300' from a tbl_teradata object. Here, only 2 rows have sampleid = 2 as the total number of rows which satisfy this condition is 3.
df1 <- td_sample(df, when_then = list("orderid < 300" = c(1, 3)))

# Print the results.
df1 %>% arrange(rowids)

# Check the number of rows of 'df1'.
td_nrow(df1)

# If with.replacement is set to TRUE, then 3 rows having sampleid = 2 will be returned.
df1 <- td_sample(df, when_then = list("orderid < 300" = c(1, 3)), with.replacement = TRUE)

# Check the number of rows of 'df1'.
td_nrow(df1)

# Example 9: Using stratified sampling with multiple conditions : 4 rows (1 row in first sample and 3 rows in second sample) when orderid < 300 and 2 rows when priority != "high". Note that only 3 rows are returned when orderid < 300.
df1 <- td_sample(df, when_then = list("orderid < 300" = c(1, 3), "priority <> 'high'" = 2))

# Print the results.
df1 %>% arrange(rowids)

# Example 10: Using 'case_else' argument for stratified sampling : 2 rows when orderid < 300 and 3 rows from the remaining rows (rows which doesn't satisfy orderid < 300).
df1 <- td_sample(df, when_then = list("orderid < 300" = 2), case_else = 3)

# Print the results.
df1 %>% arrange(rowids)
```
#### Notes
- No more than 16 samples can be requested per count or fraction list i.e. the arguments cannot take a list of elements with more than 16 samples.
- Sum of elements in the list containing fraction values should not be greater than 1 and each value should be greater than 0.
- If the list contains a float value greater than 1, e.g. `c(3, 2.4)`, then the floor value is considered for sampling i.e. first sample contains 3 rows and second sample contains 2 rows.
- A new column 'sampleid' is added to the sampled data to determine the sample set each row belongs to in the sample. If the parent tbl object already has the column 'sampleid', this column is removed in the sampled data. This case occurs when multiple or consecutive sample operations are performed. To retain the 'sampleid' columns across multiple `td_sample` operations, the column must be renamed using `mutate` function.
- If the number of samples requested exceed the number of rows available, the sample size is reduced to the number of remaining rows when the argument `with.replacement` is set to FALSE.