---
title: "Set Operations"
output:
  html_document:
    toc: true 
    toc_depth: 3
    number_sections: false
    theme: united
vignette: >
  %\VignetteIndexEntry{Set Operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 6L, tibble.print_max = 6L, digits = 3)
library(knitr)
```
# Introduction to Set Operations
The goal of this vignette is to get the user familiar with the usage of objects of class "tbl_teradata" in the following set operations provided by the dbplyr package:

- `union` (equivalent to Teradata's UNION)
- `union-all` (equivalent to Teradata's UNION-ALL)
- `intersect` (equivalent to Teradata's INTERSECT)
- `setdiff` (equivalent to Teradata's MINUS and EXCEPT)

Each set operation has a section in this vignette. Each section can be read in isolation, except where references are made to other sections or other vignettes.

#### Notes
- dbplyr does not provide set operations equivalent to INTERSECT ALL and MINUS/EXCEPT ALL.


```{r create_connection, echo=FALSE, include=FALSE}
library(dplyr)
library(dbplyr)
library(DBI)
library(tdplyr)

# Reading Vantage cluster details from td_config.cfg
config_file <- system.file("extdata", "td_config.cfg", package = "tdplyr")
config_df <- read.csv(config_file, sep = ":", stringsAsFactors = FALSE, header = TRUE)
config_list <- unlist(lapply(config_df$value, function(x) {trimws(x)}))

con <- td_create_context(host = config_list[1], uid = config_list[2], pwd = config_list[3], dType = "native")

```
Once the connection is established using **td_create_context**, you can load the data 
`mtcars` and then copy into the tables using **copy_to** and create respective objects(s) of class "tbl_teradata" using the
following statements:

```{r create_tbls, message=FALSE, warning=FALSE, results="hide"}
first <- mtcars[1:20, ]
second <- mtcars[10:32, ]

# Copy into a remote data source
df_first <- copy_to(con, first, name = "mtcars_first", row.names = TRUE)
df_second <- copy_to(con, second, name = "mtcars_second", row.names = TRUE)
```

```{r Print the data using **head** function}
# Print data.frame instead of tbl_teradata to display all rows.
print(as.data.frame(df_first))

# Print data.frame instead of tbl_teradata to display all rows.
print(as.data.frame(df_second))
```

## Union operation
`union` function combines two or more objects of class "tbl_teradata" into a single result tbl_teradata.
```{r union}
df_union <- df_first %>% union(df_second)

# Print data.frame instead of tbl_teradata to display full results.
print(as.data.frame(df_union))
```

## Union_all operation
`union_all` function combines two or more objects of class "tbl_teradata" into a single result tbl_teradata without removing duplicate rows.
```{r union_all}
df_union_all <- df_first %>% union_all(df_second)

# Print data.frame instead of tbl_teradata to display full results.
print(as.data.frame(df_union_all))
```

## Intersect operation
`intersect` function returns the data that exist in both the objects of class "tbl_teradata".
```{r intersect}
df_intersect <- df_first %>% intersect(df_second)

# Print data.frame instead of tbl_teradata to display full results.
print(as.data.frame(df_intersect))
```

## Setdiff operation
`setdiff` function returns difference of two objects of class "tbl_teradata".
```{r setdiff}
df_setdiff <- df_first %>% setdiff(df_second)

# Print data.frame instead of tbl_teradata to display full results.
print(as.data.frame(df_setdiff))
```
